Mechanized Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-01-13 paper-new/ordering.tex
rename "double poset" to "predomain"

\subsubsection{Double Posets}\label{sec:predomains}

As discussed above, there are two relations that we would like to define
in the semantics: a step-sensitive error ordering, and weak bisimilarity of computations.
%
The semantic objects that interpret our types should therefore be equipped with
two relations. We call these objects ``double posets''.
A double poset $A$ is a set with two relations: an partial order $\semlt_A$ on $A$, and
a reflexive, symmetric relation $\bisim_A$ on $A$.
We write the underling set of $A$ as $\ty{A}$.

We define morphisms of double posets as functions that preserve both
the ordering and the bisimilarity relation. Given double posets
$A$ and $B$, we write $f \colon A \monto B$ to indicate that $f$ is a morphism
from $A$ to $B$, i.e, the following hold:
(1) for all $a_1 \semlt_A a_2$, we have $f(a_1) \semlt_{B} f(a_2)$, and
(2) for all $a_1 \bisim_A a_2$, we have $f(a_1) \bisim_{B} f(a_2)$.

@@

We define an ordering on morphisms of double posets as
$f \le g$ if for all $a$ in $\ty{A_i}$, we have $f(a) \le_{A_o} g(a)$,
and similarly bisimilarity extends to morphisms via
$f \bisim g$ if for all $a$ in $\ty{A_i}$, we have $f(a) \bisim_{A_o} g(a)$.

For each type we want to represent, we define a double poset for the corresponding semantic
type. For instance, we define a double poset for natural numbers, for the
dynamic type, for functions, and for the lift monad. We
describe each of these below.

\begin{itemize}
  \item There is a double poset $\Nat$ for natural numbers, where the ordering and the
  bisimilarity relations are both equality.

  % TODO explain that there is a theta operator for posets?
  \item There is a double poset $\Dyn$ to represent the dynamic type. The underlying type
  for this double poset is defined by guarded fixpoint to be such that
  $\ty{\Dyn} \cong \mathbb{N}\, +\, \later (\ty{\Dyn} \monto \li \ty{\Dyn})$.
  This definition is valid because the occurrences of Dyn are guarded by the $\later$.
  The ordering is defined via guarded recursion by cases on the argument, using the
  ordering on $\mathbb{N}$ and the ordering on monotone functions described above.

  \item For a double poset $A$, there is a double poset $\li A$ for the ``lift'' of $A$
  using the lift monad. We use the same notation for $\li A$ when $A$ is a type
  and $A$ is a double poset, since the context should make clear which one we are referring to.
  The underling type of $\li A$ is simply $\li \ty{A}$, i.e., the lift of the underlying
  type of $A$.
  The ordering on $\li A$ is the ``lock-step error-ordering'' which we describe in
  \ref{subsec:lock-step}. The bismilarity relation is the ``weak bisimilarity''
  described in Section \ref{}

  \item For double posets $A_i$ and $A_o$, we form the double poset of monotone functions
  from $A_i$ to $A_o$, which we denote by $A_i \To A_o$.

  \item Given a double poset $A$, we can form the double poset $\later A$ whose underlying
  type is $\later \ty{A}$. We define $\tilde{x} \le_{\later A} \tilde{y}$ to be
  $\later_t (\tilde{x}_t \le_A \tilde{y}_t)$.
\end{itemize}

@@

As mentioned, the ordering on the lift of a double poset $A$ is called the
\emph{step-sensitive error-ordering} (also called ``lock-step error ordering''),
the idea being that two computations $l$ and $l'$ are related if they are in
lock-step with regard to their intensional behavior, up to $l$ erroring.
Formally, we define this ordering as follows:

@@

We also define a heterogeneous version of this ordering between the lifts of two
different double posets $A$ and $B$, parameterized by a relation $R$ between $A$ and $B$.

@@

For a double poset $A$, we define a relation on $\li A$, called ``weak bisimilarity",
written $l \bisim l'$. Intuitively, we say $l \bisim l'$ if they are equivalent
``up to delays''.
% We introduce the notation $x \sim_A y$ to mean $x \le_A y$ and $y \le_A x$.
% TODO if A is a poset, then we can just say that x = y
%
The weak bisimilarity relation is defined by guarded fixpoint as follows:

>>>>

\subsubsection{Predomains}\label{sec:predomains}

As discussed above, there are two relations that we would like to define
in the semantics: a step-sensitive error ordering, and weak bisimilarity of computations.
%
The semantic objects that interpret our types should therefore be equipped with
two relations. We call these objects ``predomains''.
A predomain $A$ is a set with two relations: an partial order $\semlt_A$ on $A$, and
a reflexive, symmetric relation $\bisim_A$ on $A$.
We write the underling set of $A$ as $\ty{A}$.

We define morphisms of predomains as functions that preserve both
the ordering and the bisimilarity relation. Given predomains
$A$ and $B$, we write $f \colon A \monto B$ to indicate that $f$ is a morphism
from $A$ to $B$, i.e, the following hold:
(1) for all $a_1 \semlt_A a_2$, we have $f(a_1) \semlt_{B} f(a_2)$, and
(2) for all $a_1 \bisim_A a_2$, we have $f(a_1) \bisim_{B} f(a_2)$.

@@

We define an ordering on morphisms of predomains as
$f \le g$ if for all $a$ in $\ty{A_i}$, we have $f(a) \le_{A_o} g(a)$,
and similarly bisimilarity extends to morphisms via
$f \bisim g$ if for all $a$ in $\ty{A_i}$, we have $f(a) \bisim_{A_o} g(a)$.

For each type we want to represent, we define a predomain for the corresponding semantic
type. For instance, we define a predomain for natural numbers, for the
dynamic type, for functions, and for the lift monad. We
describe each of these below.

\begin{itemize}
  \item There is a predomain $\Nat$ for natural numbers, where the ordering and the
  bisimilarity relations are both equality.

  % TODO explain that there is a theta operator for posets?
  \item There is a predomain $\Dyn$ to represent the dynamic type. The underlying type
  for this predomain is defined by guarded fixpoint to be such that
  $\ty{\Dyn} \cong \mathbb{N}\, +\, \later (\ty{\Dyn} \monto \li \ty{\Dyn})$.
  This definition is valid because the occurrences of Dyn are guarded by the $\later$.
  The ordering is defined via guarded recursion by cases on the argument, using the
  ordering on $\mathbb{N}$ and the ordering on monotone functions described above.

  \item For a predomain $A$, there is a predomain $\li A$ for the ``lift'' of $A$
  using the lift monad. We use the same notation for $\li A$ when $A$ is a type
  and $A$ is a predomain, since the context should make clear which one we are referring to.
  The underling type of $\li A$ is simply $\li \ty{A}$, i.e., the lift of the underlying
  type of $A$.
  The ordering on $\li A$ is the ``lock-step error-ordering'' which we describe in
  \ref{subsec:lock-step}. The bismilarity relation is the ``weak bisimilarity''
  described in Section \ref{}

  \item For predomains $A_i$ and $A_o$, we form the predomain of monotone functions
  from $A_i$ to $A_o$, which we denote by $A_i \To A_o$.

  \item Given a predomain $A$, we can form the predomain $\later A$ whose underlying
  type is $\later \ty{A}$. We define $\tilde{x} \le_{\later A} \tilde{y}$ to be
  $\later_t (\tilde{x}_t \le_A \tilde{y}_t)$.
\end{itemize}

@@

As mentioned, the ordering on the lift of a predomain $A$ is called the
\emph{step-sensitive error-ordering} (also called ``lock-step error ordering''),
the idea being that two computations $l$ and $l'$ are related if they are in
lock-step with regard to their intensional behavior, up to $l$ erroring.
Formally, we define this ordering as follows:

@@

We also define a heterogeneous version of this ordering between the lifts of two
different predomains $A$ and $B$, parameterized by a relation $R$ between $A$ and $B$.

@@

For a predomain $A$, we define a relation on $\li A$, called ``weak bisimilarity",
written $l \bisim l'$. Intuitively, we say $l \bisim l'$ if they are equivalent
``up to delays''.
% We introduce the notation $x \sim_A y$ to mean $x \le_A y$ and $y \le_A x$.
% TODO if A is a poset, then we can just say that x = y
%
The weak bisimilarity relation is defined by guarded fixpoint as follows:
